\section{Local Feature Extraction}
Beyond previously covered techniques like the \textbf{Good Features to Track} algorithm (tracked via the \textbf{Lucas-Kanade} method), 
this section introduces other local feature extraction methods, primarily the \textbf{Histogram of Oriented Gradients (HOG)} and \textbf{Scale-Invariant Feature Transform (SIFT)}.

\subsection{Histogram of Oriented Gradients (HOG)}

\subsubsection{Gradients}
Gradients are vectors pointing toward the greatest change in image intensity, making them crucial for edge detection. 
Each gradient is characterized by its \textbf{magnitude} (intensity of change) and \textbf{orientation} (direction).
It is well known that intensity and direction of gradients are one of the most important features for characterizing objects in images.

\subsubsection{Computing HOG}
\begin{itemize}
    \item Divide the image into small (rectangular or radial) cells to isolate dominant gradients.
    \item Compute the gradient magnitude and orientation for each pixel within the cell.
    \item Build a \textbf{histogram of gradient orientations} for each cell, where each pixel's contribution is weighted by its gradient magnitude.
    \item Locally normalize the histogram (e.g., via L1 or L2-norm) to account for variations in illumination and contrast.
\end{itemize}

\subsubsection{Classification using HOG}
HOG descriptors are commonly fed into classifiers, such as Support Vector Machines (SVMs), for object detection. A standard pipeline involves:
\begin{enumerate}
    \item Normalize the input image color and illumination.
    \item Compute image gradients.
    \item Build weighted histograms for each cell.
    \item Normalize the cell histograms.
    \item Concatenate the normalized histograms across the entire detection window.
    \item Feed the combined descriptor into a classifier.
\end{enumerate}

\subsubsection{The Scale Problem}
HOG is inherently not scale-invariant, which causes issues when detecting objects (like humans) at varying distances from the camera. 
This is resolved using a \textbf{multi-scale image pyramid}: the image is iteratively downsampled, and HOG features 
are extracted at each level to detect objects of different sizes.

\subsubsection{Feature Compression}
Transmitting on-device computed HOG descriptors to a central server for classification is a common requirement in edge computing, 
but the high dimensionality of these vectors can lead to severe bandwidth limitations. 
The goal of feature compression is to drastically reduce the payload size while maintaining the integrity of the data required for accurate classification.

The first step in this pipeline is \textbf{Spatial Binning}, a technique that reduces the dimensionality by grouping adjacent cells. 
This is typically optimized by applying a foveated structure: 
smaller, high-resolution cells are concentrated at the center to capture the primary object of interest, 
while larger, aggregated cells are used to summarize the background near the borders.

Finally, the binned representations are compressed using entropy encoding techniques like Huffman coding. 
By mapping the most frequent bin values to the shortest binary codes and rarer values to longer ones, the system achieves significant size reduction while preserving the exact informational content of the binned descriptors.

\subsection{Scale-Invariant Feature Transform (SIFT)}
The Scale-Invariant Feature Transform (SIFT) algorithm is designed to detect and describe highly distinctive local features that are invariant to image scale and rotation. 
While the previously discussed Good Features to Track algorithm is primarily limited to corner detection at a single scale, SIFT identifies keypoints that maintain stability across varied scales and orientations.

\subsubsection{Scale-Space Construction}
The same structure in the world appears large when the camera is close and small when far away, so detecting a feature regardless of its apparent size requires representing the image at every possible amount of blur, not just one.
This motivates the \textit{continuous scale space}: a representation in which scale is treated as a real-valued dimension $\sigma$ alongside the two spatial coordinates, smoothly sweeping from a sharp image (small $\sigma$) to a heavily blurred one (large $\sigma$).
A feature can then be located not only in space but also at its \emph{characteristic scale}, which is what ultimately makes the descriptor scale-invariant.
To achieve this, SIFT detects keypoints within this scale space rather than the original spatial domain.
The scale space is constructed by progressively convolving the image with Gaussian smoothing filters of increasing standard deviation.
Formally, the scale space function $L(x, y, \sigma)$ is defined as:
\begin{equation}
    L(x, y, \sigma) = I(x, y) * \frac{1}{2\pi\sigma^2} e^{-\frac{x^2 + y^2}{2\sigma^2}}
\end{equation}
where $I(x, y)$ is the input image and $\sigma$ is the scale parameter, which dictates the variance of the Gaussian kernel.

Since the continuous scale space cannot be stored exactly, it is sampled at discrete blur levels, organized into \textbf{octaves}, each spanning a doubling of the blur level $\sigma$.
Two distinct operations must be kept separate:
\begin{itemize}
    \item \textbf{Within an octave (fixed resolution).} Starting from a base blur $\sigma_0$, the base image is repeatedly convolved with Gaussians so that $\sigma$ grows by a constant multiplicative factor $k$ at each step ($\sigma_0,\ k\sigma_0,\ k^2\sigma_0, \dots$). An octave is therefore a \emph{stack} of images at the same resolution but with increasing blur.
    \item \textbf{Between octaves (halved resolution).} To form the next octave, the algorithm takes the image within the current octave whose blur equals $2\sigma_0$ and downsamples it by a factor of two (keeping every second pixel). After downsampling, this image's effective blur on the coarser pixel grid is back to $\sigma_0$, so the scale progression continues seamlessly.
\end{itemize}
Downsampling the $2\sigma_0$ image (rather than re-blurring the original) is an efficiency choice: a large absolute blur is far cheaper to apply on a small image than on the full-resolution one.
This between-octave step is the same Gaussian-pyramid idea used in the Lucas--Kanade pyramidal implementation; the difference is that each octave is not a single image but an entire stack spanning the intra-octave blur dimension.

\subsubsection{Extrema Detection (Difference of Gaussians)}
To efficiently locate stable keypoints across the scale space, SIFT computes the Difference of Gaussians (DoG) by subtracting adjacent blur levels within each octave:
\begin{equation}
    D(x, y, \sigma) = L(x, y, k\sigma) - L(x, y, \sigma)
\end{equation}
which provides a close approximation of the scale-normalized Laplacian of Gaussian while being cheap to compute.
A pixel in the resulting DoG volume is then compared against its 26 nearest neighbors: 8 in the current scale, 9 in the scale immediately above, and 9 in the scale immediately below.
If the pixel's value is a strict local maximum or minimum among all 26 neighbors, it is isolated as a candidate keypoint.

\subsubsection{Keypoint Refinement}
After initial detection, SIFT performs a refinement step to improve the stability and localization accuracy of candidate keypoints.
Since extrema detected in the Difference of Gaussians (DoG) volume may be affected by noise, 
SIFT refines their locations using a second-order Taylor expansion of the DoG function around the candidate keypoint:
\begin{equation}
    D(\mathbf{X}) = D + \frac{\partial D^T}{\partial \mathbf{X}}\mathbf{X}
    + \frac{1}{2}\mathbf{X}^T
    \frac{\partial^2 D}{\partial \mathbf{X}^2}\mathbf{X}
\end{equation}
where $\mathbf{X} = (x, y, \sigma)^T$ represents the spatial and scale coordinates of the keypoint.
By setting the derivative of $D(\mathbf{X})$ to zero, a more accurate estimate of the keypoint position can be obtained.
Keypoints with low contrast are discarded, as they are more sensitive to noise and less reliable for matching.

\paragraph{Edge Rejection}
The DoG function often produces strong responses along edges, which do not necessarily correspond to stable or distinctive features.
To eliminate such unstable detections, SIFT computes the \textbf{Hessian matrix} at each candidate keypoint to analyze the local curvature:
\begin{equation}
    H =
    \begin{bmatrix}
    D_{xx} & D_{xy} \\
    D_{xy} & D_{yy}
    \end{bmatrix}
\end{equation}
where $D_{xx}$, $D_{yy}$, and $D_{xy}$ denote the second-order partial derivatives of the DoG function.
Let $\alpha$ and $\beta$ represent the principal curvatures (eigenvalues) of the Hessian matrix, with ratio $r=\alpha/\beta$.
A keypoint is retained only if:
\[
    \frac{\mathrm{Tr}(H)^2}{\mathrm{Det}(H)} < \frac{(r+1)^2}{r}
\]
This criterion removes points that exhibit disproportionately large curvature in one direction, 
which are likely to correspond to edges and therefore produce unreliable feature matches.

\subsubsection{Orientation Assignment}
To achieve rotation invariance, an orientation is assigned to each refined keypoint based on the local image gradients.
For neighboring pixels around the keypoint, the gradient magnitude and orientation are computed as:
\begin{equation}
    m(x,y)=\sqrt{(L(x+1,y)-L(x-1,y))^2+(L(x,y+1)-L(x,y-1))^2}
\end{equation}
\begin{equation}
    \theta(x,y)=\tan^{-1} \left(\frac{L(x,y+1)-L(x,y-1)}{L(x+1,y)-L(x-1,y)}\right)
\end{equation}
A histogram of gradient orientations is then formed, and the dominant orientation is assigned to the keypoint.
If multiple significant peaks exist, additional keypoints may be generated with corresponding orientations.
This step ensures robustness to image rotation.

\subsubsection{Feature Descriptor}
Following orientation assignment, a local descriptor is computed around each keypoint.
A \(16 \times 16\) neighborhood centered on the keypoint and aligned with its dominant orientation is divided into \(4 \times 4\) subregions.
For each subregion, an 8-bin histogram of gradient orientations is constructed.
The resulting descriptor therefore contains:
\[
4 \times 4 \times 8 = 128
\]
elements, producing a highly distinctive feature vector that is robust to changes in illumination, rotation, and moderate scale variations.
